### 2013-06-03 doc/dissertation/document/chapters/introduction/introduction.tex
adding additional pngs to symmetric spectral analysis for communication
updates to introduction chapter

\chapter{Introduction}
\label{ch:introduction}
In many fields of engineering and physics, linear and nonlinear
problems are a primary focus of study. Recent focus on multiple
physics systems in the nuclear reactor modeling and simulation
community adds a new level of complexity to common nonlinear systems
as solution strategies change when they are coupled to other problems
\citep{u.s._department_of_energy_casl_2011}. Furthermore, a desire
for predictive simulations to enhance the safety and performance of
nuclear systems creates a need for extremely high fidelity
computations to be performed for these coupled systems as a means to
capture effects not modeled by coarser methods.

@@

We discuss in this chapter motivation for advancing Monte Carlo
solvers by providing multiphysics problems of interest in nuclear
reactor analysis. Hardware-based motivations are also provided by
considering the impact of forthcoming computing architectures. In
addition, background on the current solver techniques for multiphysics
problems and a brief comparison to the proposed methods is provided
to further motivate this work.

\section{Physics-Based Motivation}
\label{sec:physics_motivation}
Predictive modeling and simulation capability requires the combination
of high fidelity models, high performance computing hardware that can
handle the intense computational loads required by these models, and
modern algorithms for solving these problems that leverage this high
performance hardware. For nuclear reactor analysis, this predictive
capability can enable tighter design tolerances for improved thermal
performance and efficiency, higher fuel burn-up and therefore reduction
in generated waste, and high confidence in accident scenario
models. The physics that dominate these types of analysis include
neutronics, thermal hydraulics, computational fluid dynamics, and
structural mechanics.

Although solution techniques in each of these individual categories
has advanced over the last few decades and in fact leveraged modern
algorithms and computer architectures, true predictive capability for
engineered systems can only be achieved through a coupled, multiple
physics analysis where the effects of feedback between physics are
modeled. For example, consider the safety analysis of a departure from
nucleate boiling scenario in the subchannel of a nuclear fuel
assembly.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/introduction/dnb_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Multiphysics dependency analysis of departure from
      nucleate boiling.} \textit{A neutronics solution is required to
      compute power generation in the fuel pins, fluid dynamics is
      required to characterize boiling and fluid temperature and
      density, heat transfer is required to compute the fuel and
      cladding temperature, and the nuclear data modified with the
      temperature and density data. Strong coupling among the
      variables creates strong nonlinearities.}}
  \label{fig:dnb_example}
\end{figure}
When this event occurs, heat transfer is greatly reduced between the
fuel and the coolant due to the vapor layer generated by boiling,
causing the fuel center-line temperature to rapidly rise. To
characterize this boiling phenomena and how it affects fuel failure we
must consider a neutronics analysis in order to compute power
generation in the fuel pins, fluid dynamics analysis to characterize
coolant boiling, temperature, and density, solid material heat
transfer to characterize fuel and cladding temperature and heat
transfer with the coolant, and nuclear data processing to characterize
how changing material temperatures and densities changes the cross
sections needed for the neutronics calculation. As shown in
Figure~\ref{fig:dnb_example}, many couplings are required among
individual physics components in order to accurately model this
situation with each physics generating and receiving many
responses. Those variables that are very tightly coupled, such as the
temperatures generated by the fluid dynamics and heat transfer
components, will have strong nonlinearities in their behavior and
would therefore benefit from fully consistent nonlinear solution
schemes instead of fixed-point type iterations between
physics\footnote{Fixed-point iterations between physics are commonly
  referred to as Picard iterations.}. Furthermore, the space and time
scales over which these effects occur will also vary greatly.

@@

\section{Hardware-Based Motivation}
\label{sec:hardware_motivation}
As leadership class machines move towards the exascale, new algorithms
must be developed that leverage their strengths and adapt to their
shortcomings. Basic research is required now to advance methods in
time for these new machines to become operational. Organized work is
already moving forward in this area with the Department of Energy's
Advanced Scientific Computing Research office specifically allocating
funding for the next several years to research resilient solver
technologies for exascale facilities
\citep{u.s._department_of_energy_resilient_2012}. Based on the
language in this call for proposals, we can identify key issues for
which a set of robust, massively parallel Monte Carlo solvers could
provide a solution. As machines begin to operate at hundreds of
petaflops peak performance and beyond, trends toward reduced energy
consumption will require incredibly high levels of concurrency to
achieve the desired computation rates. Furthermore, this drop in
power consumption will mean increased pressure on memory as memory per
node is expected to stagnate while cores per node is expected to
increase. As the number of cores increases, their clock speed is
expected to stagnate or even decrease to further reduce power
consumption and manufacturing costs.

@@

\section{Research Outline}
\label{sec:research_outline}
For some time, the particle transport community has been utilizing
Monte Carlo methods for the solution of transport problems
\citep{lewis_computational_1993}. The partial differential equation
(PDE) community has focused on various deterministic methods for
solutions to linear problems \citep{saad_iterative_2003,
  kelley_iterative_1995}. In between these two areas are a not widely
known group of Monte Carlo methods for solving sparse linear systems
\citep{forsythe_matrix_1950, hammersley_monte_1964,
  halton_sequential_1962, halton_sequential_1994}. In recent years,
these methods have been further developed for radiation transport
problems in the form of Monte Carlo Synthetic-Acceleration (MCSA)
\citep{evans_monte_2009, evans_monte_2012} but have yet to be applied
to more general sparse linear systems commonly generated by the
computational physics community. Compared to other methods in this
regime, MCSA offers three attractive qualities; (1) the linear problem
operator need not be symmetric or positive-definite, thereby reducing
preconditioning complexity, (2) the stochastic nature of the solution
method provides a natural solution to the issue of resiliency, and (3)
is amenable to parallelization using modern methods developed by the
transport community \citep{wagner_hybrid_2010}. The development of
MCSA as a general linear solver and the development of a parallel MCSA
method will be new and unique features of this work, providing a
framework with which other issues such as resiliency may be addressed
in the future.

In addition to linear solver advancements, nonlinear solvers may also
benefit from a general and parallel MCSA scheme. In the nuclear
engineering community, nonlinear problems are often addressed by
either linearizing the problem or building a segregated scheme and
using traditionally iterative or direct methods to solve the resulting
system \citep{pletcher_computational_1997}. In the mathematics
community, various Newton methods have been popular
\citep{kelley_iterative_1995}. Recently, Jacobian-Free Newton-Krylov
(JFNK) schemes \citep{knoll_jacobian-free_2004} have been utilized in
multiple physics architectures and advanced single physics codes
\citep{gaston_parallel_2009}. The benefits of JFNK schemes are that
the Jacobian is never formed, simplifying the implementation, and a
Krylov solver is leveraged (typically GMRES or Conjugate Gradient),
providing excellent convergence properties for well-conditioned and
well-scaled systems. However, there are two potential drawbacks to
these methods for high fidelity predictive simulations: (1) the
Jacobian is approximated by a first-order differencing method on the
order of machine precision such that this error can grow beyond that
of those in a fine-grained system \citep{kelley_iterative_1995} and
(2) for systems that are not symmetric positive-definite (which will
be the case for most multiphysics systems and certainly for most
preconditioned systems) the Krylov subspace generated by the GMRES
solver may become prohibitively large
\citep{knoll_newton-krylov_1995}. To address these issues, this thesis
develops a new and novel method for nonlinear systems based on the
MCSA method.

@@

This preliminary report outlines in Chapter~\ref{ch:linear_problem}
the conventional methods used in practice for solving linear problems
to provide a mathematical basis upon which to build new algorithms
that aim to solve some of the aforementioned issues. Parallel schemes
for conventional methods are also provided for background and
understanding of how current methods may or may not map to future
hardware. In addition, some components of their parallel
implementations may be applied to stochastic methods development. In
Chapter~\ref{ch:stochastic_methods}, Monte Carlo algorithms for
solving linear systems and new parallel strategies will be outlined in
full with links made to past work and their potential for offering
improvements to the multiphysics analysis community. From these
stochastic methods, a new nonlinear method is developed in
Chapter~\ref{ch:nonlinear_problem} and compared to conventional
methods for solving nonlinear problems.

>>>>

\chapter{Introduction}
\label{ch:introduction}
In nearly all high-fidelity nuclear reactor simulations, linear and
nonlinear transport problems are a primary focus of study. Recent
focus on multiple physics systems in the nuclear reactor modeling and
simulation community adds a new level of complexity to common linear
and nonlinear systems as solution strategies change when they are
coupled to other problems
\citep{u.s._department_of_energy_casl_2011}. Furthermore, a desire for
predictive simulations to enhance the safety and performance of
nuclear systems creates a need for extremely high fidelity
computations to be performed for these coupled transport systems as a
means to capture effects not modeled by coarser methods.

@@

We discuss in this chapter physics-based motivation for advancing
Monte Carlo solvers and studying their application to transport
systems by providing problems of interest in nuclear reactor
analysis. Hardware-based motivations are also provided by considering
the impact of forthcoming computing architectures. In addition,
background on the current solver techniques for multiphysics problems
and a brief comparison to the proposed methods is provided to further
motivate this work. The organization of the document is then reviewed
followed by an explanation of the key outcomes of this work and where
in the document results and data supporting those outcomes can be
located.

%%---------------------------------------------------------------------------%%
\section{Physics-Based Motivation}
\label{sec:physics_motivation}
Predictive modeling and simulation capability requires the combination
of high fidelity models, high performance computing hardware that can
handle the intense computational loads required by these models, and
modern algorithms for solving these problems that leverage this high
performance hardware. For nuclear reactor analysis, this predictive
capability can enable tighter design tolerances for improved thermal
performance and efficiency, higher fuel burn-up and therefore reduction
in generated waste, and high confidence in accident scenario
models. The physics that dominate these types of analysis include
neutronics, thermal hydraulics, computational fluid dynamics, and
structural mechanics.

Although solution techniques in each of these individual categories
has advanced over the last few decades and in fact leveraged modern
algorithms and computer architectures, true predictive capability for
engineered systems can only be achieved through a coupled, multiple
physics analysis where the effects of feedback between physics are
modeled. For example, consider the safety analysis of a departure from
nucleate boiling scenario in the subchannel of a nuclear fuel
assembly.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/introduction/dnb_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Multiphysics dependency analysis of departure from
      nucleate boiling.} \textit{A neutronics solution is required to
      compute power generation in the fuel pins, fluid dynamics is
      required to characterize boiling and fluid temperature and
      density, heat transfer is required to compute the fuel and
      cladding temperature, and the nuclear data modified with the
      temperature and density data. Strong coupling among the
      variables creates strong nonlinearities.}}
  \label{fig:dnb_example}
\end{figure}
When this event occurs, heat transfer is greatly reduced between the
fuel and the coolant due to the vapor layer generated by boiling,
causing the fuel center-line temperature to rapidly rise. To
characterize this boiling phenomena and how it affects fuel failure we
must consider a neutronics analysis in order to compute power
generation in the fuel pins, fluid dynamics analysis to characterize
coolant boiling, temperature, and density, solid material heat
transfer to characterize fuel and cladding temperature and heat
transfer with the coolant, and nuclear data processing to characterize
how changing material temperatures and densities changes the cross
sections needed for the neutronics calculation. As shown in
Figure~\ref{fig:dnb_example}, many couplings are required among
individual physics components in order to accurately model this
situation with each physics generating and receiving many
responses. Those variables that are very tightly coupled, such as the
temperatures generated by the fluid dynamics and heat transfer
components, will have strong nonlinearities in their behavior and
would therefore benefit from fully consistent nonlinear solution
schemes instead of fixed-point type iterations between
physics\footnote{Fixed-point iterations between physics are commonly
  referred to as Picard iterations.}. Furthermore, the space and time
scales over which these effects occur will also vary greatly.

@@

%%---------------------------------------------------------------------------%%
\section{Hardware-Based Motivation}
\label{sec:hardware_motivation}
As leadership class machines move towards the exascale, new algorithms
must be developed that leverage their strengths and adapt to their
shortcomings. Basic research is required now to advance methods in
time for these new machines to become operational. Organized work is
already moving forward in this area with the Department of Energy's
Advanced Scientific Computing Research office specifically allocating
funding for the next several years to research resilient solver
technologies for exascale facilities
\citep{u.s._department_of_energy_resilient_2012}. Based on the
language in this call for proposals, we can identify key issues for
which a set of robust, massively parallel Monte Carlo solvers could
provide a solution. As machines begin to operate at hundreds of
petaflops peak performance and beyond, trends toward reduced energy
consumption will require incredibly high levels of concurrency to
achieve the desired computation rates. Furthermore, this drop in
power consumption will mean increased pressure on memory as memory per
node is expected to stagnate while cores per node is expected to
increase. As the number of cores increases, their clock speed is
expected to stagnate or even decrease to further reduce power
consumption and manufacturing costs.

@@

%%---------------------------------------------------------------------------%%
\section{Research Outline}
\label{sec:research_outline}
For some time, the particle transport community has been utilizing
Monte Carlo methods for the solution of transport problems
\citep{lewis_computational_1993}. The partial differential equation
(PDE) community has focused on various deterministic methods for
solutions to linear problems \citep{saad_iterative_2003,
  kelley_iterative_1995}. In between these two areas are a not widely
known group of Monte Carlo methods for solving sparse linear systems
\citep{forsythe_matrix_1950, hammersley_monte_1964,
  halton_sequential_1962, halton_sequential_1994}. In recent years,
these methods have been further developed for radiation transport
problems in the form of Monte Carlo Synthetic-Acceleration (MCSA)
\citep{evans_monte_2009, evans_monte_2012} but have yet to be applied
to more general sparse linear systems commonly generated by the
computational physics community. Compared to other methods in this
regime, MCSA offers three attractive qualities; (1) the linear problem
operator need not be symmetric or positive-definite, thereby reducing
preconditioning complexity, (2) the stochastic nature of the solution
method provides a natural solution to the issue of resiliency, and (3)
is amenable to parallelization using modern methods developed by the
transport community \citep{wagner_hybrid_2010}. The development of
MCSA as a general linear solver and the development of a parallel MCSA
method are new and unique features of this work, providing a
framework with which other issues such as resiliency may be addressed
in the future.

In addition to linear solver advancements, nonlinear solvers may also
benefit from a general and parallel MCSA scheme. In the nuclear
engineering community, nonlinear problems are often addressed by
either linearizing the problem or building a segregated scheme and
using traditionally iterative or direct methods to solve the resulting
system \citep{pletcher_computational_1997}. In the mathematics
community, various Newton methods have been popular
\citep{kelley_iterative_1995}. Recently, Jacobian-Free Newton-Krylov
(JFNK) schemes \citep{knoll_jacobian-free_2004} have been utilized in
multiple physics architectures and advanced single physics codes
\citep{gaston_parallel_2009}. The benefits of JFNK schemes are that
the Jacobian is never formed, simplifying the implementation, and a
Krylov solver is leveraged (typically GMRES or Conjugate Gradient),
providing excellent convergence properties for well-conditioned and
well-scaled systems. However, there are two potential drawbacks to
these methods for high fidelity predictive simulations: (1) the
Jacobian is approximated by a first-order differencing method on the
order of machine precision such that this error can grow beyond that
of those in a fine-grained system \citep{kelley_iterative_1995} and
(2) for systems that are not symmetric positive-definite (which will
be the case for most multiphysics systems and certainly for most
preconditioned systems) the Krylov subspace generated by the GMRES
solver may become prohibitively large
\citep{knoll_newton-krylov_1995}. To address these issues, this work
develops a new and novel method for nonlinear systems based on the
MCSA method.

@@

To present these new developments this document is arranged in the
following manner. First, in Chapter~\ref{ch:stochastic_methods}, the
fundamentals of the Monte Carlo method for linear systems are
presented. Using this background on Monte Carlo, synthetic
acceleration methods are then presented and analyzed using a simple
model transport problem. Next, in Chapter~\ref{ch:spn_equations}, the
Monte Carlo methods developed in Chapter~\ref{ch:stochastic_methods}
are applied to the neutron transport problem. Specifically, the $SP_N$
form of the Boltzmann neutron transport equation is developed and
solved using the Monte Carlo methods. Using a difficult fission
reactor fuel assembly criticality calculation to drive research and
development for these Monte Carlo methods, several important issues
regarding Monte Carlo solver applicability and performance will be
discussed along with solutions to these problems. Using these
solutions, the Monte Carlo methods are then verified against modern
solution techniques for the $SP_N$ equations in order verify
correctness for the fuel assembly criticality calculation. In
Chapter~\ref{ch:nonlinear_problem}, the Monte Carlo methods are
applied to the Navier-Stokes equations as a model nonlinear energy and
momentum transport system. Additional difficulties that arise when
using Monte Carlo methods in these types of systems will be presented
along with the subsequently developed solutions. In
Chapter~\ref{ch:parallel_methods}, the Monte Carlo methods are
parallelized using modern reactor physics techniques for particle
transport with a full parallel scaling analysis provided using a
leadership-class computing facility. Finally, the work is summarized
in Chapter~\ref{ch:conclusion} and topics of future work derived from
the results of this research presented.

### 2013-06-03 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
updates to introduction chapter

%%---------------------------------------------------------------------------%%
\section{MCSA Comparison to Conventional Methods}
\label{sec:spn_comparison}
Through the numerical studies presented up to this point in this
chapter, we have developed a set of preconditioning techniques that
permit MCSA to be used with the $SP_N$ form of the neutron transport
equation and applied to a difficult fuel assembly criticality
problem. Along with these preconditioning techniques, a set a varying
MCSA parameters were analyzed to find those that yielded the best
performance for this particular problem. In addition, several steps
were taken to mitigate the dense composite linear operators that arise
when performing explicit MCSA preconditioning in order to achieve
convergence. Using these results, we compare MCSA directly to
conventional linear solvers that would be typically used to solve the
$SP_N$ form of the transport equation as presented here. In
particular, we will use a pair of production Krylov solvers with which
we will compare direct numerical results in order to verify MCSA
solutions and iterative and CPU timing performance in order to gauge
feasibility and identify areas that require further research.

\subsection{MCSA Verification}
\label{subsec:spn_mcsa_verification}
In order to verify the correctness of the MCSA method and
implementation used for this work, MCSA solutions to the fuel assembly
problem with 1, 2, and 4 energy groups using $SP_1$ discretization
will be directly compared over all group and cell fluxes computed by
converging the criticality problem using both BiCGStab and GMRES
production implementations from the Trilinos scientific computing
library Aztec \citep{heroux_overview_2005}. In addition to comparing
the cell and group fluxes, the number of eigenvalue iterations
required to converge the problem along with the eigenvalue itself will
be compared as a means of verification.

@@

\subsection{Performance Results}
\label{subsec:spn_comparison_results}
Using the same set of solver parameters used for verification, MCSA
will be directly compared to the production Krylov solvers in terms of
both iterative performance and CPU timing. In addition to the
parameters used for the verification analysis, artificial absorption
was also introduced with a value of 0.2 for the MCSA calculations that
used the collision esitmator as CPU timing performance was improved
without a loss in iterative performance. Adding this absorption did
not change the numerical results of the
calculation.

>>>>

%%---------------------------------------------------------------------------%%
\section{MCSA Verification}
\label{sec:spn_mcsa_verification}
Through the numerical studies presented up to this point in this
chapter, we have developed a set of preconditioning techniques that
permit MCSA to be used with the $SP_N$ form of the neutron transport
equation and applied to a difficult fuel assembly criticality
problem. Along with these preconditioning techniques, a set a varying
MCSA parameters were analyzed to find those that yielded the best
performance for this particular problem. In addition, several steps
were taken to mitigate the dense composite linear operators that arise
when performing explicit MCSA preconditioning in order to achieve
convergence. Using these results, we compare MCSA directly to
conventional linear solvers that would be typically used to solve the
$SP_N$ form of the transport equation as presented here. In
particular, we will use a pair of production Krylov solvers with which
we will compare direct numerical results in order to verify MCSA
solutions and iterative and CPU timing performance in order to gauge
feasibility and identify areas that require further research.

In order to verify the correctness of the MCSA method and
implementation used for this work, MCSA solutions to the fuel assembly
problem with 1, 2, and 4 energy groups using $SP_1$ discretization
will be directly compared over all group and cell fluxes computed by
converging the criticality problem using both BiCGStab and GMRES
production implementations from the Trilinos scientific computing
library Aztec \citep{heroux_overview_2005}. In addition to comparing
the cell and group fluxes, the number of eigenvalue iterations
required to converge the problem along with the eigenvalue itself will
be compared as a means of verification.

@@

%%---------------------------------------------------------------------------%%
\section{MCSA Performance Comparison to Conventional Methods}
\label{sec:spn_comparison}
Using the same set of solver parameters used for verification, MCSA
will be directly compared to the production Krylov solvers in terms of
both iterative performance and CPU timing. In addition to the
parameters used for the verification analysis, artificial absorption
was also introduced with a value of 0.2 for the MCSA calculations that
used the collision esitmator as CPU timing performance was improved
without a loss in iterative performance. Adding this absorption did
not change the numerical results of the calculation.
